---
title: Capstone Data Cleaning
author: Siobhan Oberg
date: '2021-08-04'
slug: capstone-data-cleaning
categories:
  - Coding
  - R
  - Capstone
tags:
  - Coding
  - Background
  - Capstone
  - Cleaning
---

Loading Libraries
```{r, echo = FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(assertive)
library(stringr)
library(lubridate)
library(VIM)
library(maps)
library(ggmap)
library(ggplot2)  
library(rgdal)  
library(raster)
library(sf)
library(leaflet)
```

This code reads in the csv file that has the data in it. You need to enter the pathway to the file so that R can find it. On my computer, the csv file is located in my Downloads folder, which is in my user account on the computer, asoberg, which is in the Users folder.
```{r data}
salem_trees <- read_csv("/Users/asoberg/Downloads/Trees_lat_and_long.csv")
```

The glimpse function gives tou a overview of your data, telling you how many rows, columns, the class of each variable, and the first few observations of each column. Though the variables are listed vertically here, the variables are actually arranged horizontally when you look at the actual data frame.
```{r trees}
glimpse(salem_trees)
```

I wanted to see a count of all the different values for most of the variables. There are a few reasons why.
-See how many different values there are, and to get a count of how many times each value occured.
-See if there were any missing values, to get a count of how many missing values per column.
-See if there were any values that could be compressed into one uniform value, ie replacing a miss spelling of a word with the correct spelling to have be counted towards the correct value count.

To do that, I used the table function, which requires the dataset name and the column name I am interested in. The $ is used to extract a particular column.

I have notes that I wrote to myself about the columns below the table function for each column. 
```{r pressure, echo=FALSE}
table(salem_trees$'Planting Area Type')
```
Most common planting area type is planting strip between sidewalk and road.
```{r}
table(salem_trees$'Planting Area Width in Feet')
```
Need to average out the area width and assign that to the null values.

```{r}
table(salem_trees$'Overhead Utilities')
```
Don't know what these are.

```{r}
table(salem_trees$'Tree Present')
```
Hmmm, were the trees stolen? removed? 

```{r}
table(salem_trees$'Tree Genus/Species')
```
Outcome variable

```{r}
table(salem_trees$Cultivar)
```
Related to outcome variable.

```{r}
table(salem_trees$'Exact DBH in Inches')
```
A feature to maybe help determine species, means diameter at breast height.
```{r}
table(salem_trees$'Tree Health')
```
Only 1 in excellent health. 

```{r}
table(salem_trees$Topped)
```
Top part of tree is cut off, can cause decay, cavaties, hollows.
```{r}
table(salem_trees$Cavities)
```
What are cavities? are they just holes in a tree?
```{r}
table(salem_trees$Fungus)
```
Good that only 72 trees have fungus, not all fungi growing on your tree are harmful; some do not affect the tree at all while others are even beneficial. Maybe a trend will be revealed after mapping the trees? Is there a specific area of Salem where trees are have fungi?

```{r}
table(salem_trees$'Large Dead Limbs')
```
There are almost no comments, so just remove that whole column.

```{r}
table(salem_trees$Address)
```
Just the street name for the most part, don't know that that would be useful, esp when I also have the lat and long.

```{r}
table(salem_trees$Ward)
```
Don't know what this means.

```{r}
table(salem_trees$UNITTYPE)
```
There are only 408 park trees, 10118 right of way trees, yet they're all city of salem trees?

```{r}
table(salem_trees$PLANTDATE)
```
Not sure this adds much value, why are they all planted on january first? (doubtful this is true) Can I assume the ones with null were planted before 2005?

```{r}
table(salem_trees$STATUS)
```
What? No clue what an active tree is and most of them are null anyway.

```{r}
table(salem_trees$OWNEDBY)
```
All owned by city of salem parks, and I'm sure all maintained by them too (the answer is yes). 

```{r}
table(salem_trees$PWGRID)
```
Remove this column.

```{r}
table(salem_trees$MEMORIALNAME)
```
Remove this column.

```{r}
table(salem_trees$PARK)
```
Majority of trees are not in parks. Leads me to believe that each park maintains their own trees.

```{r}
table(salem_trees$'Planting Notes')
```
This will need a lot of cleaning, if I even want to use planting notes, as most are null
and then lat and long. Might need to change lat and long class.

I decided I don't want to deal with the planting notes, not super important, might be a post for later.

Here I am subsetting the data set into a smaller one, removing the columns I am not interested in. This does not decrease the number of observations, just the number of variables.
```{r}
less_trees = subset(salem_trees, select = -c(Shape, `Overhead Utilities`, PWGRID, MEMORIALNAME, Comments, Ward, UNITTYPE, OBJECTID, MAINTAINEDBY, OWNEDBY, STATUS, Cultivar, PLANTDATE, `Planting Notes`))
```

I will say this data is suprisingly clean, there aren't different variations of the same thing (no misspellings, weird capitalizations, etc) except in a column I'm not even going to use. The majority of time spent in data science projects, whether they be making machine learning models and artificial intelligence, or creating forecasts, or most anything else, is pent on data cleaning. A lot of the data you find in the wild will require more text cleaning, and standarization of values, replacing of missing dat and more. Even though I said this data is in better condition than most, I do need to transform some variables in numeric from character, create numeric categories, streamline column titles, impute some values, make all alphabetical characters lowercase, etc. 

Making column titles cleaner

I want to make it easier and more efficient to type the column titles. So I fully change the column titles. 

Here, I use the names function to change the names of a few columns to shorter values, as well as to remove the spaces in the column titles. Having spaces in the column titles adds complications to typing out column names. When there are spaces in column names, you need to put '' around the name, which causes R to not suggest the column names when typing them out. I use the <- to write over a value in this example. The <- assigns the the functions, values, etc after it to the value before it.
```{r}
names(less_trees)[names(less_trees) == "Planting Area Type"] <- "pa_type"
names(less_trees)[names(less_trees) == "Planting Area Width in Feet"] <- "pa_width_ft"
names(less_trees)[names(less_trees) == "Tree Present"] <- "tree_present"
names(less_trees)[names(less_trees) == "Tree Genus/Species"] <- "genus_species"
names(less_trees)[names(less_trees) == "Exact DBH in Inches"] <- "DBH_in"
names(less_trees)[names(less_trees) == "Tree Health"] <- "health"
names(less_trees)[names(less_trees) == "Topped"] <- "top"
names(less_trees)[names(less_trees) == "Cavities"] <- "cav"
names(less_trees)[names(less_trees) == "Fungus"] <- "fungus"
names(less_trees)[names(less_trees) == "Large Dead Limbs"] <- "lg_dead_limbs"
names(less_trees)[names(less_trees) == "Address"] <- "address"
names(less_trees)[names(less_trees) == "PARK"] <- "park"
names(less_trees)[names(less_trees) == "LATITUDE"] <- "lat"
names(less_trees)[names(less_trees) == "LONGITUDE"] <- "long"
```

Lowering characters

This bit of cleanup causes all alphabetical characters to be lower case. I use the function tolower to do this. This use of the <- applys a function to all the values in a column. 
```{r}
less_trees$pa_type<- tolower(less_trees$pa_type)
less_trees$pa_width_ft<- tolower(less_trees$pa_width_ft)
less_trees$tree_present<- tolower(less_trees$tree_present)
less_trees$genus_species<- tolower(less_trees$genus_species)
less_trees$DBH_in<- tolower(less_trees$DBH_in)
less_trees$health<- tolower(less_trees$health)
less_trees$top<- tolower(less_trees$top)
less_trees$cav<- tolower(less_trees$cav)
less_trees$fungus<- tolower(less_trees$fungus)
less_trees$lg_dead_limbs<- tolower(less_trees$lg_dead_limbs)
less_trees$address<- tolower(less_trees$address)
less_trees$pARK<- tolower(less_trees$park)
```

Changing Null to NA

In order to change a from one data type to another, they all need to actual values or NA. R will be angry if for instance there observations that are alphabetical characters when the rest of the observations are numbers. So i replace all instances of <null> with NA. R treats NA as a blank cell. 
```{r}
less_trees[less_trees == "<null>"] <- NA
```

Visualizing missing data

While we're on the topic of NAs (missing data), why not vizualize it?
```{r}
aggr(less_trees, numbers = TRUE, prop = c(TRUE, FALSE))
```
All this red doesn't look fantastic, but it is useful to be able to see it all laid out like this. This helps me determine if there are other variables that I should remove from the data set, and if there are variables that it makes sense to impute.


Changing data types from character to numeric

In order for me to impute anything, I need to change the data classes of a few columns. Specifically, the columns that need to be numeric.
```{r}
less_trees$pa_width_ft <- as.numeric(less_trees$pa_width_ft)
less_trees$DBH_in <- as.numeric(less_trees$DBH_in)
```

Now I can replace the missing values in the pa_width and DBH columns.

Imputing pa width

Here, I just replace missing values with the avgerage planting area width. 
```{r}
less_trees$pa_width_ft[is.na(less_trees$pa_width_ft)] <- mean(less_trees$pa_width_ft, na.rm=TRUE)
```

Grouping and imputing DBH

Imputing DBH is a little bit harder. I want a more exact mean, so I decided to replace the missing values by tree species with the mean of that tree species DBH. I do this by using the aggregate function which groups the data by the values in the fifth column(in this example), and then calculate the mean.
```{r}
aggregate(less_trees[, 5], list(less_trees$genus_species), mean)
```

I then group by the genus species column and if there is a NA value I replace it with the species average.
```{r}
less_trees <- less_trees %>%
    group_by(genus_species) %>%
    mutate(DBH_in = ifelse(is.na(DBH_in), mean(DBH_in, na.rm = TRUE), DBH_in))
```

```{r}
summary(less_trees)
```
This tells us there are still a few DBH that are NA. So I am just going to impute them as 0.
```{r}
less_trees$DBH_in[is.na(less_trees$DBH_in)] <- 0
```

Replacing NA genus species with unknown

I want to visualize how many trees that we don't know the species of, and we can only do that by assigning a value to those missing points.
```{r}
less_trees[c("genus_species")][is.na(less_trees[c("genus_species")])] <- "unknown"
```

Replacing NA pa type with unknown

Same reason for replacing genus species.
```{r}
less_trees[c("pa_type")][is.na(less_trees[c("pa_type")])] <- "unknown"
```

Basic mapping of points

Right now, longitude and latitude are acting as the x and y coordinates of a graph, and so we can plot the trees as points on a graph.
```{r}
prelim_plot <- ggplot(less_trees, aes(x = long, y = lat, 
    colour = park)) +
    geom_point()
prelim_plot
```
This graph shows all the trees managed and owned by the City of Salem. If you're familiar with the Salem geography, you can see the outline of the Willamette River that cuts Salem in half, just by the large area that has no points. 


Fixing lat and long
You may have noticed that the latitude and longitude are not actual latitude and longitude values, and so I need to convert them to the correct coordinates. To do this I first subset the less tress dataset into just the lat and long columns. 
```{r}
salem_lat_long = subset(less_trees, select = c(lat, long))
```

I then identify the coordinate referencing system from ArcGIS, so that in the next bit of code I can transform that CRS to a more conventional lat and long system.

```{r}
salem_lat_long <- salem_lat_long %>% 
    st_as_sf(coords = c("long", "lat"), crs = 2913)
salem_lat_long
```
This CRS is the NAD83(HARN) / Oregon North (ft) CRS and I am converting to the WGS84 CRS.

```{r}
salem_lat_long <- st_transform(salem_lat_long, "+proj=longlat +ellps=WGS84 +datum=WGS84")
```
Now the two columns have been combined into one column and I need to split them. This is where the str_split_fixed function comes in. 

I tell the function to split the string in the geometry column into 2 separate columns, at the comma between the two series of numbers.
```{r}
salem_lat_long <- str_split_fixed(salem_lat_long$geometry, ", ", 2)
```

I then need to convert the matrix into data frame in order to subset and remove unnecessary characters.
```{r}
salem_lat_long <- as.data.frame(salem_lat_long)
```

Longitude

I subset the salem_lat_long in to salem_long and remove unnecessary characters, covnert to a data frame again, covnert the data class into numeric, and then set up a new data frame with id numbers for each longitude value, and bind the id_long to the salem_long, so that I can rejoin the salem_long to the less_trees data frame.
```{r}
#long
salem_long <- subset(salem_lat_long, select = c(V1))
salem_long <- str_replace_all(salem_lat_long$V1, "[c(]", "")
salem_long <- as.data.frame(salem_long)
salem_long$salem_long <- as.numeric(as.character(salem_long$salem_long))
id_long <- rownames(salem_long)
salem_long <- cbind(id_long=id_long, salem_long)
```

Latitude
Same process as the longitude.
```{r}
salem_lat <- subset(salem_lat_long, select = c(V2))
salem_lat <- str_replace_all(salem_lat_long$V2, "[)]", "")
salem_lat <- as.data.frame(salem_lat)
salem_lat$salem_lat <- as.numeric(as.character(salem_lat$salem_lat))
id_lat <- rownames(salem_lat)
salem_lat <- cbind(id_lat=id_lat, salem_lat)
```

Rejoining
I convert the latitude an longitude columns into numeric classes. I create an id column for less_trees as I did for salem_lat and salem_long. I then join the salem_lat and salem_long columns using the full_join function. 
```{r}
less_trees$lat <- as.numeric(less_trees$lat)
less_trees$long <- as.numeric(less_trees$long)
id_salem <- rownames(less_trees)
less_trees <- cbind(id_salem=id_salem, less_trees)
less_trees <- less_trees %>% full_join(salem_lat, by = c("id_salem" = "id_lat"))
less_trees <- less_trees %>% full_join(salem_long, by = c("id_salem" = "id_long"))
```

I remove the innaccurate lat and long, as well as id_salem, as they are no longer needed. 
```{r}
less_trees = subset(less_trees, select = -c(lat, long, id_salem))
```

Renaming salem_lat, salem_long, just to clean it up.
```{r}
names(less_trees)[names(less_trees) == "salem_lat"] <- "latitude"
names(less_trees)[names(less_trees) == "salem_long"] <- "longitude"
```

Quick calculation of Salem's overall Biodiversity
```{r}
198/(10527 - 1499)
```
198 is the number of different species in Salem, excluding the trees which species we don't know. Then I am dividing by the number of total trees, minus the amount that are unknown species. 

0.02 is the biodiversity index, which is pretty low. 

I create 2 new datasets for the visualizations I want to create, and the models I want to create. 
Data viz Dataset
```{r}
tree_viz = subset(less_trees, select = c(genus_species, latitude, longitude, fungus, top, health, park, health, lg_dead_limbs))
```

Modelling Dataset
```{r}
tree_model = subset(less_trees, select = c(genus_species, cav, top, health, pa_type,pa_width_ft, fungus, lg_dead_limbs, DBH_in))
```

Lastly I export these 2 dataframes  and less_trees as csv's so I can use them in other parts of the project. 
```{r}
write.csv(tree_viz, "/Users/asoberg/Downloads/tree_viz.csv")
```

```{r}
write.csv(tree_model, "/Users/asoberg/Downloads/tree_model.csv")
```

```{r}
write.csv(less_trees, "/Users/asoberg/Downloads/less_trees.csv")
```

